---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "README-")
```

## Clustering and Classification in R: Osama Mahmoud and Berthold Lausen

Notes, practicals and R code for the up coming [course](https://www.essex.ac.uk/iads/events/summer-school.aspx) at 
the University of Essex Big Data and Analytics Summer School, September 2016.

  * PDF version of course [slides](https://raw.githubusercontent.com/Osmahmoud/essexBigdata/master/slides.pdf).
  
## Installing the `essexBigdata` package

The practicals and associated R commands, use a variety of R packages. The easiest way of installing all dependencies is to install `drat` first, as follows:

```{r eval=FALSE}
install.packages("drat")
```

Then use `drat` to add a new repository
```{r, eval=FALSE}
drat::addRepo("Osmahmoud")
```

The way for installing packages should now work using
```{r eval=FALSE}
install.packages("essexBigdata")
```

## Usage of practicals

The `essexBigdata` package includes a number of practical files to guide participants to applications of clustering and classification techniques using R. 

  * Practical1 - Basic concepts
  
```{r eval=FALSE}
vignette("practical1", package = "essexBigdata")
```

  * Practical2 - Cluster analysis

```{r eval=FALSE}
vignette("practical2", package = "essexBigdata")
```

## Further reading

  - **Everitt, B., Hothorn, T. (2011)**, [An Introduction to Applied Multivariate Analysis with R](http://link.springer.com/book/10.1007%2F978-1-4419-9650-3) - Springer-Verlag.

  - **Everitt, B., Landau, S., Leese, M., Stahl, D. (2011)**, [Cluster Analysis](http://eu.wiley.com/WileyCDA/WileyTitle/productCd-EHEP002266.html), Wiley.

  - **James, G., Witten, D., Hastie, T., Tibshirani, R. (2013)**, [An Introduction to Statistical Learning with Applications in R](http://www.springer.com/gb/book/9781461471370), Springer-Verlag.
  
  - **Mahmoud, O. (2015)**, [Improving Statistical Learning within Functional Genomic Experiments by means of Feature Selection](https://www.researchgate.net/publication/304704707_Improving_Statistical_Learning_within_Functional_Genomic_Experiments_by_means_of_Feature_Selection), PhD thesis. University of Essex. Available from: [The Albert Sloman Library](http://serlib0.essex.ac.uk/search~S5?/aMahmoud%2C+Osama/amahmoud+osama/-3%2C-1%2C0%2CB/frameset&FF=amahmoud+osama&1%2C1%2C).
