---
title: "Practical 1"
author: "Osama Mahmoud"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{practical1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Basics in Statistics

To get you familiar with underlying concepts of cluster analysis (unsupervised learning) and classification (supervised learning), we would use the `measure` data set, which is included within the `essexBigdata` package, to illustrate some basic concepts in statistics e.g. covariance, correlation and Euclidean distance.

### Covariance

The covariance of two random variables $X_{i}$ and $X_{j}$, denoted by $\sigma_{i,j}$, is a measure of linear dependence and defined by:

$$\sigma_{i,j} = Cov(X_{i}, X_{j}) = E[(X_{i} - \mu_{i})(X_{j} - \mu{j})],$$

with $\mu_{i} = E(X_{i})$ and $\mu_{j} = E(X_{j})$.

Note that, when $i = j$ we observe that the covariance $Cov(X_{i}, X_{i})$ is the variance of the
random variable $X_{i}$ which is denoted by $\sigma_{i}^2$ and can be defined as:

$$\sigma_{i}^2 = Cov(X_{i}, X_{i}) = E(X_{i} - \mu_{i})^2.$$

Larger absolute values of the covariance imply a higher degree of linear dependence between the two variables.

#### Covariance matrix

For q-dimensional multivariate random vectors $X = (X_{1}, ... , X_{q})$, we have $q$ variances $\sigma_{i}^2$ for $i = 1, ... , q$ and $\frac{q(q-1)}{2}$ covariances. The variances and covariances can be arranged in $q × q$ symmetric matrix $\Sigma$ as follows:

$$\Sigma = 
\begin{pmatrix}
  \sigma_{1}^2 & \sigma_{1,2} & \cdots & \sigma_{1,q} \\
  \sigma_{2,1} & \sigma_{2}^2 & \cdots & \sigma_{2,q} \\
  \vdots  & \vdots  & \ddots & \vdots  \\
  \sigma_{q,1} & \sigma_{q,2} & \cdots & \sigma_{q}^2
\end{pmatrix}$$

For a sample of $q$-dimensional observations, $x_{1}, ... , x_{n}$ with $x_{i} \in \mathbb{\mathbb{R}}^{q}$ for $i = 1, ... , n$, we can estimate the covariance matrix $\Sigma$ by estimating corresponding variances and covariances from the given sample.

We can use `R` to calculate the covariance matrix for the `measure` data set. To begin with, load the `measure` data

``` {r}
#data(measure, package = "essexBigdata")
```

Next, you can show the first six observations and details of the `measure` data sets

```{r}
#head(measure)
```
